Monte Carlo estimation of the probability of causal contacts between communicating civilisations

### 2020-05-30 submit_v03/manuscript.tex
last changes

%The authors can define any packages after the \documentclass{CSL} command.
\usepackage{amsmath} %for dealing with mathematics,
\usepackage{amsthm} %for dealing with theorem environments,
\usepackage{cite} %%%for dealing with citations
\usepackage{hyperref} %for linking the cross references
\usepackage{graphics} %for dealing with figures.
\usepackage{algorithmic} %for describing algorithms
\usepackage{subfig} %for getting the subfigures e.g., "Figure 1a and 1b" etc.
\usepackage{url} %It provides better support for handling and breaking URLs.

%\usepackage{natbib}
\usepackage{boites}

%\usepackage{graphicx}
%\usepackage{float}
%\usepackage{psfig}
%\usepackage{amsfonts,amssymb}
%\usepackage{color}
%\usepackage{textcomp}
%\usepackage[T1]{fontenc}
%\usepackage{aecompl}
%\usepackage{times}
%\usepackage{txfonts}
%\usepackage[english]{babel}
%\usepackage[utf8x]{inputenc}
%\usepackage{amsmath}
%\usepackage{caption}
%\usepackage{siunitx}

@@

\newcommand{\ceti}{C3N}
\newcommand{\cetis}{C3Ns}
\newcommand{\hs}{\textsc{hearsay}}

@@

The Drake equation  \citep{drake_intelligent_1962} provides a truly
helpful educated guess, a rational set of lenses --the factors in the
equation-- through which to look at future contacts with
technologically advanced civilizations in the Milky Way. 
%
The equation quantifies the number of civilizations from whom we might
receive an electromagnetic signal, using a collection of factors that
have been extensively discussed in the literature and whose estimates
are revised continually. 
%
A comprehensive review and analysis of each term of the equation are
presented in \citet{vakoch_drake_2015}.
%
Optimistic estimates from the Drake equation contrast with the
so-called Fermi paradox, which states the apparent contradiction between the
expected abundance of life in the Galaxy and the lack of evidence for
it \citep[e.g. ][]{hart_explanation_1975, brin_great_1983,
barlow_galactic_2012, forgan_galactic_2016, anchordoqui_is_2017,
Sotos_biotechnology_2019, carroll_nellemback_fermi_2019}.
%
There are many propositions aimed at solving this paradox, which make
use of statistical \citep{solomonides_probabilistic_2016, horvat_calculating_2007,
maccone_statistical_2015} or stochastic approaches
\citep{forgan_numerical_2009, bloetscher_using_2019,
glade_stochastic_2011, forgan_numerical_2010}.
%
Regarding the Drake equation, analytical interpretations
\citep{prantzos_joint_2013, smith_broadcasting_2009} or reformulations
\citep[][and references therein]{burchell_whither_2006} have also been
proposed. 
%
The absence of detections of extraterrestrial intelligent signals
could be explained by
astrophysical phenomena that makes life difficult to develop
\citep{annis_astrophysical_1999}.
%
Besides the possible scarcity of life,
alternative scenarios have also been discussed
\citep{barlow_galactic_2013, lampton_information_2013,
conway_three_2018, forgan_galactic_2017}.
%
The large distances in the Galaxy and the likely limited lifetime of
civilizations may play an important role in determining how difficult it
would be to obtain evidence for other inhabited worlds.
%
The analysis of these scenarios is difficult due to the lack of data
about the hypothetical extraterrestrial intelligences.
%
Indeed, as \citep{tarter_search_2001} pointed out, according to our
current technical capabilities for the search of extraterrestrial
intelligence (SETI), we have not received any signal yet.
%
The absence of detections has also motivated alternative ideas for new
SETI strategies \citep{forgan_exoplanet_2017, balbi_impact_2018,
loeb_eavesdropping_2006, maccone_KLT_2010, tarter_advancing_2009,
enriquez_breakthrough_2017, loeb_relative_2016, maccone_SETI_2011,
lingam_relative_2019, wright_theGsearch_2015, maccone_SETI_2013,
maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

The discussion about the problem of the unknown
abundance of civilisations in the Galaxy has been organised around the
factors in the Drake equation \citep{hinkel_interdisciplinary_2019}.
%
However, the uncertainties in these factors, specially the ones
representing biological processes, make it
less suitable to a formal study with the purpose of defining
searching strategies or
computing the estimated number of extraterrestrial intelligences.
%
A number of studies propose alternative formalisms for the estimation of the
likelihood of detecting intelligent signals from space.
%
\citet{prantzos_joint_2013}, for example, proposes a unified framework for a joint
analysis of the Drake equation and the Fermi paradox, concluding that
for sufficiently long-lived civilisations, colonisation
is the most promising strategy to find other life forms.
%
\citet{haqq-misra_drake_2017} discuss the dependence of the Drake
equation parameters on the spectral type of the host stars and the
time since the Galaxy formed, and examine trajectories for the
emergence of communicative civilisations.
%
Some modifications to the original idea of the Drake equation have
been proposed, in order to
integrate a temporal structure, to
reformulate it as a stochastic process or to propose alternative
probabilistic expressions.
%
Temporal aspects of the distribution of communicating civilisations
and their contacts have been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011}, as
well as efforts on considering the stochastic nature of the Drake
equation \citep{glade_stochastic_2011}.
%
On this topic a number of works consider
numerical simulations \citep{forgan_evaluating_2015,
vukotic_grandeur_2016, murante_simulating_2015, forgan_numerical_2009,
forgan_galactic_2017, ramirez_new_2017}.
%
Although this approach does not rely on values that can be measured
from observations
(like the fraction of stars with planets), 
it depends on the definition of
unknown or uncertain parameters required to carry out
a the simulations.
%
\citet{grimaldi_signal_2017} considers a statistical model for the
probability of the Earth contacting other intelligent civilisations,
taking into account the finite lifetime of signal emitters, and based
on the fractional volume occupied by all signals reaching our planet.

Several authors raise the distinction between causal contacts and
actual contacts.
%
In the first case, the determinations of metrics for the likelihood of
a contact are independent of considerations about the technological
resources or the implicit coordination to decipher intelligent
messages.
%
In a recent work, \citet{balbi_impact_2018} uses a statistical model to
analyse the occurrence of causal contacts between civilisations in the
Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilisations that
might be in causal contact with an observer on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasises the lack of temporal
structure in the Drake equation and, in particular, the limitations
of this expression to estimate the required timescale of a SETI
program to succeed in the detection of intelligent signals.

@@

A notable example of statistical modeling is the work by
\citet{balbi_impact_2018}, who investigates the chance of
communicating civilisations making causal contact within a volume
surrounding the location of the Earth.
%
The author argues that the causal contact requirement involves mainly
the distance between civilisations, their lifespan and their times of
appearance.
%
This is important since the time the light takes to travel across the
Galaxy might be much lesser than the lifetime of the emitter.
%
\citet{balbi_impact_2018} fixes the total number of civilisations and
explores the parametric space that comprises three parameters, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilization.
%
Each of these three variables are drawn from a random distribution.
%
The distribution of the distances results from a uniform model for the
positions of civilizations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores exponential and truncated Gaussian distributions,
while for the distribution of the lifespans chooses an exponential
distribution.
%
It is important to consider that the estimation of the number of
communicating civilisations vary with the choice of the statistical
distribution for the time of appearance, as shown by
\citet{balbi_impact_2018}.
%
For all analysed distributions, the author concludes that, the
fraction of emitters that are listened is low if they are spreaded in
time and with limited lifetimes.
%
In another numerical approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account the neighbour cells.
%
The authors implement this model to a network of cells which represent
life complexity on a 2-dimensional region resembling the Galactic
Habitable Zone (GHZ), an annular ring set between a minimal radius of
6 kpc and a peak radius of 10 kpc.
%
These simulations represent the spread if intelligence as an
implementation of panspermia theories.
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
several Monte Carlo simulations and analyse ensemble-averaged results.
%
Their work aims at analysing the evolution of life, and although it
does not account for the network of causal contacts among
technological civilisations, it offers a tool to think SETI searches
from a novel point of view.

@@

This study is conceived as an introduction to a simple probabilistic
model and a numerical exploration of its parameter space.
%
With these tools, we address the problem of the temporal and spatial structure of
the distribution of communicating civilisations.
%
This approach does not require asumptions about, e.g., the origin of life, the development of
intelligence or the formation of habitable planets according to stellar type.
%
Instead, we assume monoparametric function families to model the appearance of
points in the disk of the Galaxy and over the time, which we call
\textit{nodes}.
%
These nodes represent the locations of ideal intelligent agents that are able
to receive and emit signals with perfect efficiency in all directions.
%
Then we analyse the network of causally connected nodes limited to have a
maximum separation representing the maximum distance a signal can travel with
an intensity above a fixed threshold.
%
A node is causally connected to other node if it is within their light cone.
%
We adopt as a definition that a light cone is the region in spacetime within
the surface generated by the light emanating from a given point in space in a
given period of time.
%
For example, two nodes separated by 100 light years which appear with a
difference in time of 80 years and last 50 years each, will be causally
connected for 30 years.
%
This connection is not, though, bidirectional.
%
In fact, the second node will see the first one for 30 years, but the first
node will never acknowledge the existence of the second.
%
This simple tought experiment exemplify the importance of the time variable to
analyse the structure of connections.
%
The model can be described with three parameters, namely, the mean separation
between the appearance of nodes, the mean lifetime of the nodes, and the
maximum separation to allow an effective causal contact.
%
The model presented here comprises the stochastic networks of constrained
causally connected nodes, together with the three parameters and additional
hypotheses.
%
Hereafter, we refer to this model as the SC3Net model, standing for
''Stochastic Constrained Causally Connected Network``.

@@

This paper is organized as follows.
%
In Sec.~\ref{S_methods} we introduce the methods and discuss the candidate
distributions for the statistical aspects of the times involved in the
communication process.
%
We present our results in Sec~\ref{S_results}, with special emphasis on the
statistical distributions of the duration of causal contacts in one or both
directions and the distribution of time intervals for the waiting of the first
contact.
%
This quantities are considered as a function of the three simulation
parameters.
%
In Sec.~\ref{S_discussion} we discuss our results and future research
directions.

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is the sphere centred on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the region
$R$.
%
The surface of last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The receiver node E$_1$ can listen signals from the emitter node E$_2$, from
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$. 
%
} \label{F_scheme}
%
\end{figure}

@@

In what follows, we outline the experimental setup adopted to estimate the
probabilities of causal contacts and several derived quantities.
%
This is done in terms of three independent parameters, namely, the mean time
span between the appearance of consecutive \cetis{}, $\left<\tau_a\right>$,
their mean lifetime $\left<\tau_s\right>$, and the maximum distance a signal
can be detected by another \ceti{} ($D_{\rm max}$).
%
Intuitively, the probability of the existence of causal contacts between pairs
of \cetis{} would be larger for smaller $\tau_a$ parameters, higher $\tau_s$ or
higher $D_{\rm max}$ parameter values.
%
We also propose theoretical distribution functions for both the lifespans
($\tau_s$) and the number of \cetis{} per unit time
\citep{maccone_evolution_2014, Sotos_biotechnology_2019}.
%
The later is related to the time span between the appearance of consecutive
\cetis{} (since when $\left<\tau_a\right>$ is shorter, it produces a greater
density of \cetis{}).
%
The analytical expresions for these distributions are set to a fixed law, as
discussed in Sec.~\ref{SS_PDF_shape}.

\ffn{1} Space—time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and
space axes are years and light years, respectively.
%
Panel (a) of the represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Fig.~\ref{F_scheme}.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon.
%
This polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ``Awakening'' events ($A_1$ and $A_2$).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ``Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ``Blackout'' events, presented as $B_1$ and
$B_2$.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ``C'' and ``B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

In our simulations, we assume the simplest configuration for the
growth of the sphere of first contact.
%
In particular, we do not consider the
possibility of stellar colonisation
\citep[e.g.][]{newman_galactic_1981, walters_interstellar_1980,
starling_virulence_2013, barlow_galactic_2012, jeong_large_2000,
maccone_mathematical_2011}.
%
We also assume that communication is
equally likely in all directions, i.e., we assume isotropic
communication in all cases.
%
Different communication efficiencies or
detection methods are straightforward to carry out in the simulations
for more detailed and complex scenarios.
%
This approach, however, is
beyond this work because it obscures the experimental setup and make
the results less clear.
%
The assumptions we accept imply that the
results are independent of whether intelligent agents are organic or
artificial.
%
Moreover, the causes of the limited lifetime of a
civilisation can be natural (astrophysical phenomena), caused by auto
destruction or by external factors, to name a few.
%
However, we assume
that these events are sufficiently numerous in the Galaxy so that a
statistical model is plausible. The failure of this hypothesis would
imply extreme values for the parameters that represent the density of
the nodes (i.e., $\left< \tau_a \right>$).

 \setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{cllll}
\hline
   \multicolumn{2}{l}{independent variables (parameter space)}
   &min. value&max. value&Nbins\\
%\hline
%   $\tau_{a}$ & Mean temporal separation between consecutive awakenings 
%	      & 500~a & 40500~a & 21\\ 
%	    & & 5$\cdot$10$^5$~a & 10$^6$~a & 51\\ 
%   $\tau_{s}$ & Mean lifetime of a \ceti{}
%	      & 500~a & 40500~a & 21\\ 
%	    & & 5$\cdot$10$^5$~a & 10$^6$~a & 51\\ 
%	D$_{\rm max}$ & Maximum reach of a message  & \multicolumn{3}{c}{100 pc, 500 pc, 1 kpc, 5 kpc, 10 kpc} \\
\hline

>>>>

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{graphics}
\usepackage{algorithmic}
\usepackage{subfig}
\usepackage{url}

\usepackage{boites}

@@

\newcommand{\ceti}{node}
\newcommand{\cetis}{nodes}
\newcommand{\hs}{\textsc{hearsay}}

@@

The Drake equation  \citep{drake_intelligent_1962} provides a truly
helpful educated guess, a rational set of lenses --the factors in the
equation-- through which to look at future contacts with
technologically advanced civilisations in the Milky Way. 
%
The equation quantifies the number of civilisations from whom we might
receive an electromagnetic signal, using a collection of factors that
have been extensively discussed in the literature and whose estimated
values are revised continually. 
%
A comprehensive review and an analysis of each term of the equation are
presented in \citet{vakoch_drake_2015}.
%
Optimistic estimates from the Drake equation contrast with the
so-called Fermi paradox, which states the apparent contradiction between the
expected abundance of life in the Galaxy and the lack of evidence for
it \citep[e.g. ][]{hart_explanation_1975, brin_great_1983,
barlow_galactic_2012, forgan_galactic_2016, anchordoqui_is_2017,
Sotos_biotechnology_2019, carroll_nellemback_fermi_2019}.
%
There are many propositions aimed at solving this paradox, which make
use of statistical \citep{solomonides_probabilistic_2016, horvat_calculating_2007,
maccone_statistical_2015} or stochastic approaches
\citep{forgan_numerical_2009, bloetscher_using_2019,
glade_stochastic_2011, forgan_numerical_2010}.
%
Regarding the Drake equation, analytical interpretations
\citep{prantzos_joint_2013, smith_broadcasting_2009} or reformulations
\citep[][and references therein]{burchell_whither_2006} have also been
proposed. 
%
The absence of detections of extraterrestrial intelligent signals
could be explained by
astrophysical phenomena that makes life difficult to develop
\citep{annis_astrophysical_1999}.
%
Besides the possible scarcity of life,
alternative scenarios have also been discussed
\citep{barlow_galactic_2013, lampton_information_2013,
conway_three_2018, forgan_galactic_2017}.
%
The large distances in the Galaxy and the likely limited lifetime of
civilisations may play an important role in determining how difficult it
would be to obtain evidence for other inhabited worlds.
%
The analysis of these scenarios is difficult due to the lack of data
about the hypothetical extraterrestrial intelligences.
%
Indeed, as \citet{tarter_search_2001} pointed out, according to our
current technical capabilities for the search of extraterrestrial
intelligence (SETI), we have not received any signal yet.
%
The absence of detections has also motivated alternative ideas for new
SETI strategies \citep{forgan_exoplanet_2017, balbi_impact_2018,
loeb_eavesdropping_2006, maccone_KLT_2010, tarter_advancing_2009,
enriquez_breakthrough_2017, loeb_relative_2016, maccone_SETI_2011,
lingam_relative_2019, wright_theGsearch_2015, maccone_SETI_2013,
maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

The discussion about the problem of the unknown abundance of
civilisations in the Galaxy has been organised around the factors in
the Drake equation \citep{hinkel_interdisciplinary_2019}.
%
However, the uncertainties in these factors, specially the ones
representing biological processes, make it less suitable to a formal
study with the purpose of defining searching strategies or computing
the estimated number of extraterrestrial intelligences.
%
A number of studies propose alternative formalisms for the estimation
of the likelihood of detecting intelligent signals from space.
%
\citet{prantzos_joint_2013}, for example, proposes a unified framework
for a joint analysis of the Drake equation and the Fermi paradox,
concluding that for sufficiently long-lived civilisations,
colonisation is the most promising strategy to find other life forms.
%
\citet{haqq-misra_drake_2017} discuss the dependence of the Drake
equation parameters on the spectral type of the host stars and the
time since the Galaxy formed, and examine trajectories for the
emergence of communicative civilisations.
%
Some modifications to the original idea of the Drake equation have
been proposed, in order to integrate a temporal structure, to
reformulate it as a stochastic process or to propose alternative
probabilistic expressions.
%
Temporal aspects of the distribution of communicating civilisations
and their contacts have been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011}, as
well as efforts on considering the stochastic nature of the Drake
equation \citep{glade_stochastic_2011}.

Several authors raise the distinction between causal contacts and
actual contacts.
%
In the first case, the determinations of metrics for the likelihood of
a contact are independent of considerations about the technological
resources or the implicit coordination to decipher intelligent
messages.
%
In a recent work, \citet{balbi_impact_2018} uses a statistical model to
analyse the occurrence of causal contacts between civilisations in the
Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilisations that
might be in causal contact with an observer on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasises the lack of temporal
structure in the Drake equation and, in particular, the limitations
of this expression to estimate the required timescale of a SETI
program to succeed in the detection of intelligent signals.
% 
\citet{balbi_impact_2018} also investigates the chance of
communicating civilisations making causal contact within a volume
surrounding the location of the Earth.
%
The author argues that the causal contact requirement involves mainly
the distance between civilisations, their lifespan and their times of
appearance.
%
This is important since the time the light takes to travel across the
Galaxy might be much lesser than the lifetime of the emitter.
%
\citet{balbi_impact_2018} fixes the total number of civilisations and
explores the parametric space that comprises three variables, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilisations.
%
Each of these three variables are drawn from a random distribution.
%
The distances are drawn from a uniform model 
for the
positions of civilisations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores exponential and truncated Gaussian functions,
while for the lifespans chooses an exponential behaviour.
%
It is important to point out that the estimation of the number of
communicating civilisations vary with the choice of the statistical
model for the time of appearance, as shown by
\citet{balbi_impact_2018}.
%
For all analysed distributions, the author concludes that the
fraction of emitters that are listened is low if they are spreaded in
time and with limited lifetimes.
%
An analytical explanation of these concepts are presented 
in \citet{grimaldi_signal_2017}, who considers a statistical model for the
probability of the Earth contacting other intelligent civilisations,
taking into account the finite lifetime of signal emitters, and based
on the fractional volume occupied by all signals reaching our planet.

@@

On this topic a number of works consider numerical simulations
\citep{forgan_evaluating_2015, vukotic_grandeur_2016,
murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017,
ramirez_new_2017}.
%
Although this approach does not rely on values that can be measured
from observations (like the fraction of stars with planets), it
depends on the definition of unknown or uncertain parameters required
to carry out the simulations.
%
In another numerical approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account the neighbour cells.
%
The authors implement this model to a network of cells which represent
life complexity on a two--dimensional region resembling the Galactic
Habitable Zone (GHZ), an annular ring set between a minimal radius of
6~kpc and a peak radius of 10~kpc.
%
These simulations represent the spread of intelligence as an
implementation of panspermia theories.
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
Monte Carlo simulations and analyse ensemble-averaged results.
%
Their work aims at analysing the evolution of life, and although it
does not account for the network of causal contacts among
technological civilisations, it offers a tool to think SETI searches
from a novel point of view.

@@

This study is conceived as an introduction to a simple probabilistic
model and a numerical exploration of its parameter space.
%
With these tools, we address the problem of the temporal and spatial
structure of the distribution of communicating civilisations.
%
This approach does not require asumptions about, e.g., the origin of
life, the development of intelligence or the formation of habitable
planets according to stellar type.
%
Instead, we assume monoparametric function families to model the
appearance of points in the disk of the Galaxy and over the time,
which we call \textit{nodes}.
%
These nodes represent the locations of ideal intelligent agents that
are able to receive and emit signals with perfect efficiency in all
directions.
%
Then we analyse the network of causally connected nodes limited to
have a maximum separation representing the maximum distance a signal
can travel with an intensity above a fixed threshold.
%
A node is causally connected to other node if it is within their light
cone.
%
We adopt as a definition that a light cone is the region in spacetime
within the surface generated by the light emanating from a given point
in space in a given period of time.
%
For example, two nodes separated by 100 light years which appear with
a difference in time of 80 years and last 50 years each, will be
causally connected for 30 years.
%
This connection is not, though, bidirectional.
%
In fact, the second node will see the first one for 30 years, but the
first node will never acknowledge the existence of the second.
%
An scheme of this example is shown if Fig.~\ref{F_example}.
%
This simple tought experiment exemplify the importance of the time
variable to analyse the structure of connections.
%
The model can be described with three parameters, namely, the mean
separation between the appearance of nodes, the mean lifetime of the
nodes, and the maximum separation to allow an effective causal
contact.
%
The model presented here comprises the stochastic networks of
constrained causally connected nodes, together with the three
parameters and additional hypotheses.
%
Hereafter, we refer to this model as the SC3Net model, standing for
''Stochastic Constrained Causally Connected Network``.

\begin{figure}[h]
%
\centering
%
\includegraphics[width=\columnwidth]{F_example.pdf}
%
\caption{Example of a configuration of an emitter and a receiver 
   to produce a causal contact.
   %
   The observers are separated by 100~light years and appear with a
   difference in time of 80~years. One observer is active between
   $t=$0~yr and $t=$50~yr.  The other observer is also active for
   50~yr, between $t=$80~yr and $t=$130~yr. 
   %
   The causal connection occurs between $t=$100~yr and $t=$130~yr
   %
   Meaningfull times are marked with vertical dashed lines.
   }
%
\label{F_example}
%
\end{figure}

@@

This paper is organized as follows.
%
In Sec.~\ref{S_methods} we introduce the methods and discuss the
candidate distributions for the statistical aspects of the times
involved in the communication process.
%
We present our results in Sec~\ref{S_results}, with special emphasis
on the statistical distributions of the duration of causal contacts in
and the distribution of time intervals of waiting for the first contact.
%
This quantities are considered as a function of the three simulation
parameters.
%
In Sec.~\ref{S_discussion} we discuss our results and future research
directions.

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node.
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is the sphere centred on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
This sphere is represented by the left triangle of the region
$R$.
%
The surface of last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The node E$_1$ can listen signals from the node E$_2$, since
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the node E$_2$ is in causal contact from the
   node E$_1$, in the time interval (C$_{21}$, B$_{21}$).
%
} \label{F_scheme}
%
\end{figure}

@@

In what follows, we outline the experimental setup adopted to estimate the
probabilities of causal contacts and several derived quantities.
%
This is done in terms of three independent parameters, namely, the mean time
span between the appearance of consecutive \cetis{}, $\left<\tau_{\rm
a}\right>$,
their mean lifetime $\left<\tau_{\rm s}\right>$, and the maximum distance a signal
can be detected by another \ceti{} ($D_{\rm max}$).
%
Intuitively, the probability of the existence of causal contacts between pairs
of \cetis{} would be larger for smaller $\tau_{\rm a}$ parameters,
higher $\tau_{\rm s}$ or
higher $D_{\rm max}$ parameter values.
%
We also propose theoretical distribution functions for both the lifespans
($\tau_{\rm s}$) and the number of \cetis{} per unit time
\citep{maccone_evolution_2014, Sotos_biotechnology_2019}.
%
The later is related to the time span between the appearance of consecutive
\cetis{} (since when $\left<\tau_{\rm a}\right>$ is shorter, it produces a greater
density of \cetis{}).
%
The analytical expresions for these distributions are set to a fixed law, as
discussed in Sec.~\ref{SS_PDF_shape}.

\ffn{1} Space--time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and
space axes are years and light years, respectively.
%
Panel (a) of the represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Fig.~\ref{F_scheme}.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon.
%
This polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ``Awakening'' events ($A_1$ and $A_2$).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ``Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ``Blackout'' events, presented as $B_1$ and
$B_2$.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ``C'' and ``B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

In our simulations, we assume the simplest configuration for the
growth of the sphere of first contact.
%
In particular, we do not consider the
possibility of stellar colonisation
\citep[e.g.][]{newman_galactic_1981, walters_interstellar_1980,
starling_virulence_2013, barlow_galactic_2012, jeong_large_2000,
maccone_mathematical_2011}.
%
We also assume that communication is
equally likely in all directions, i.e., we assume isotropic
communication in all cases.
%
Different communication efficiencies or
detection methods are straightforward to carry out in the simulations
for more detailed and complex scenarios.
%
This approach, however, is
beyond this work because it obscures the experimental setup and make
the results less clear.
%
The assumptions we accept imply that the
results are independent of whether intelligent agents are organic or
artificial.
%
Moreover, the causes of the limited lifetime of a
civilisation can be natural (astrophysical phenomena), caused by auto
destruction or by external factors, to name a few.
%
However, we assume
that these events are sufficiently numerous in the Galaxy so that a
statistical model is plausible. The failure of this hypothesis would
imply extreme values for the parameters that represent the density of
the nodes (i.e., $\left< \tau_{\rm a} \right>$).
